Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-22 paper-filmat.tex
Prepared for double-blind review
Down to two XXX's

\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
\author{Ingo Blechschmidt}
\address{Universität Augsburg \\
Institut für Mathematik \\
Universitätsstr. 14 \\
86159 Augsburg, Germany}
\email{ingo.blechschmidt@math.uni-augsburg.de}

@@

\newextarrow{\xbigtoto}{{20}{20}{20}{20}}{\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\Zar}{Zar}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\effectively}{ef{}fectively\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

@@

\begin{abstract}
  XXX
\end{abstract}

\maketitle
\thispagestyle{empty}

@@

\begin{figure}
  \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
  \tikzstyle{title} = [fill=mypurple, text=white]

@@

\item We explicate how toposes provide distinct lenses through which the
usual mathematical objects of the standard topos can be viewed.
\end{enumerate}
% XXX check everywhere

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.
% XXX: mention Olivia?
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
% XXX contingent: To some extent, what the commonly agreed-upon rules of
% mathematics are is dependent on contingent historical factors. It is
% conceivable, for instance, that the intuitionists ...
\end{enumerate}
We invite further research on these connections.

@@

\bigskip
\paragraph{Acknowledgments} We are grateful to XXX for many invaluable
discussions shaping this work and to XXX for their careful criticism of earlier
drafts. XXX conference organizer and participants
% Todd Lehman for creating parts of Figure~1

@@

Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
so-called \emph{Kripke--Joyal translation rules}. For instance, the rule for
translating conjunction reads
\[ \E \models (\alpha \wedge \beta) \qquad\text{iff}\qquad
  \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
The remaining translation rules are more
involved~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
the case of a general topos~$\E$, but we will state them in the next sections
for several specific toposes.

@@

\subsection{The logic of toposes} By their definition as special kinds of
categories, toposes are merely algebraic structures not unlike groups or vector
spaces. Hence we need to argue why we picture toposes as mathematical universes
while we do not elevate other kinds of algebraic structures in the same way.
For us, this usage is justified by the following metatheorem:

@@

@@

\bigskip
\begin{center}
\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  % FUTURE: Countable axiom of choice
  \bottomrule
\end{tabular}
\end{center}

@@

\bigskip
\dotparagraph{``Any natural number is prime or not.''} Even without knowing what
a prime number is, one can safely judge this statement to be true in
the standard topos, since it is just an instance of the law of excluded middle.

By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
amounts to stating that there is a Turing machine which, when given a natural
number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
not. Such a Turing machine indeed exists -- writing such a program is often a
first exercise in programming courses. Hence the statement is also true in the
\effective topos, but for the nontrivial reason that primality can be
algorithmically tested.

\bigskip
\dotparagraph{``There are infinitely many primes.''} A first-order formalization
of this statement is ``for any natural number~$n$, there is a prime
number~$p$ which is greater than~$n$'', and is known to be true in the standard
topos by any of the many proofs of this fact.

Its external meaning when interpreted in the \effective topos is that there exists
a Turing machine~$M$ which, when given a natural number~$n$ as input, terminates with a
prime number~$p > n$ as output. Such a Turing machine exists, hence the
statement is true in the \effective topos.\footnote{More precisely, the
machine~$M$ should also output the description of a Turing machine which
witnesses that~$p$ is prime and that~$p > n$. However, the statement~``$p$ is prime and~$p > n$''
is~$\neg\neg$-stable (even decidable), and for those statements witnesses are
redundant.}

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

A rigorous proof that such a machine~$M$ does not exist reduces its assumed
existence to the undecidability of the halting problem.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is computable.''}
The preceding examples give the impression that what is true in the
\effective topos is solely a subset of what is true in the standard topos. The
example of this subsection, the so-called \emph{formal Church--Turing thesis},
shows that the relation between the two toposes is more nuanced.

@@

The usual proof sets up the halting function~$h : \NN \to \NN$, defined using
the case distinction
\[ h : n \mapsto \begin{cases}
  1, & \text{if the $n$-th Turing machine terminates}, \\
  0, & \text{if the $n$-th Turing machine does not terminate},
\end{cases} \]
and proceeds to show that~$h$ is not computable. However, in the \effective
topos, this definition does not give rise to a total function from~$\NN$
to~$\NN$. The actual domain is the subset of those natural numbers~$n$
for which the~$n$-th Turing machine
terminates or does not terminate. Assuming the law of excluded middle, this is
a trivial condition; but intuitionistically, it is not.

\bigskip
\dotparagraph{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the sign and Heaviside functions
being prominent counterexamples. In the \effective topos, this statement is
true and independently due to
Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
straightforward (a textbook reference
is~\cite[Thm.~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
explanation is as follows.

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, when given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
such approximations. Hence a function such as the sign function, for which
even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
precision in the input~$x$, do not exist in the \effective topos.

\bigskip
\dotparagraph{``Markov's principle holds.''} Markov's principle is the
following statement:
\begin{equation}\label{eq:markov}\tag{MP}
  \forall f \? \NN^\NN\_ \bigl((\neg\neg\exists n\?\NN\_ f(n) = 0)
  \Longrightarrow \exists n\?\NN\_ f(n) = 0\bigr).
\end{equation}
It is an instance of the law of double negation elimination and hence trivially
true in the standard topos, at least if we subscribe to classical logic on the
meta level. A useful consequence of Markov's principle is that Turing machines
which do not run forever (that is, which do \emph{not not} terminate) actually
terminate; this follows by applying Markov's principle to the function~$f : \NN
\to \NN$ where~$f(n)$ is zero or one depending on whether a given Turing
machine has terminated within~$n$ computational steps or not.

@@

As a consequence, the input given to machine~$S_F$ is entirely uninformative
and~$S_F$ cannot make direct computational use of it. But its existence ensures that an
\emph{unbounded search} will not fail (and hence succeed, by an appeal to Markov's
principle on the meta level): The machine~$S_F$ can simulate~$F$ to
compute the values~$f(0), f(1), f(2), \ldots$ in turn, and stop
with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.

\bigskip
\dotparagraph{``Heyting arithmetic is categorical.''} In addition to the standard
model~$\NN$, the standard topos contains uncountably many nonstandard models of
Peano arithmetic (at least if we assume a classical meta level). By a theorem
of Benno van den Berg and Jaap van Oosten, the situation
is quite different in the \effective topos~\cite{berg-oosten:arithmetic}:
\begin{enumerate}
\item Heyting arithmetic, the intuitionistic cousin of Peano arithmetic, has
exactly one model up to isomorphism, namely~$\NN$.
\item In fact, even the finitely axiomatizable subsystem of Heyting arithmetic
where the induction scheme is restricted to~$\Sigma_1$-formulas has exactly one
model up to isomorphism, again~$\NN$. As a consequence, Heyting arithmetic is
finitely axiomatizable.
\item Peano arithmetic is ``quasi-inconsistent'' in that it does not have any
models, for any model of Peano arithmetic would also be a model of Heyting
arithmetic, but the only model of Heyting arithmetic is~$\NN$ and~$\NN$ does
not validate the theorem ``any Turing machine terminates or does not
terminate'' of Peano arithmetic.
\end{enumerate}
As a consequence, Gödel's completeness theorem fails in the \effective topos:
In the \effective topos, Peano arithmetic is consistent (because it is
equiconsistent to Heyting arithmetic, which has a model) but does not have a
model.

The reason for~(1) is related to the fact that no nonstandard model of
Peano arithmetic in the standard topos is computable~\cite{tennenbaum:models}.

@@

In the realizability topos constructed using infinite-time Turing machines, the full
law of excluded middle still fails, but some instances which are wrong in the
\effective topos do hold in this topos. For instance, the instance ``any
function~$\NN \to \NN$ is the zero function or not'' does: Its external meaning
is that there is an infinite-time Turing machine~$M$ which, when given the
description of an infinite-time Turing machine~$F$ computing a function~$f :
\NN \to \NN$ as input, terminates (at some ordinal time step) with a correct
judgment of whether~$f$ is the zero function or not. Such a machine~$M$ indeed
exists: It simply has to simulate~$F$ on all inputs~$0,1,\ldots$ in order and
check whether one of the resulting function values is not zero. This
will require a transfinite amount of time (not least because simulating~$F$ on
just one input might require a transfinite amount of time), but as an
infinite-time Turing machine,~$M$ is capable of carrying out this procedure.

@@

% XXX introduction...

@@

@@

\subsection{Real numbers}
As detailed in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
actually a Turing machine computing arbitrarily-good consistent rational
approximations. A similarly drastic shift in meaning, though in an orthogonal
direction, occurs with~$\Sh(X)$. What~$\Sh(X)$ believes to be a (Dedekind) real
number~$a$ is actually a continuous family of real numbers on~$X$, that is, a
continuous function~$a : X \to \RR$.

@@

\subsection{Real functions} Let~$(f_x)_{x \in X}$ be a continuous family of
continuous real-valued functions; that is, each of the individual
functions~$f_x : \RR \to \RR$ should be continuous and moreover the map~$\RR
\times X \to \RR, (a,x) \mapsto f_x(a)$ should be continuous.
From the point of view of~$\Sh(X)$, this family looks like a single continuous
function~$f : \RR \to \RR$.

@@

From these observations we can deduce that the intermediate value theorem of
undergraduate calculus does in general not hold in~$\Sh(X)$ and hence does not allow
for an intuitionistic proof. The intermediate value theorem states: ``If~$f : \RR \to \RR$
is a continuous function such that~$f(-1) < 0$ and~$f(1) > 0$, there exists a
number~$a$ such that~$f(a) = 0$.'' The external meaning of this statement is
that in any continuous family~$(f_x)_x$ of continuous functions with~$f_x(-1) <
0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible to pick zeros of
the family in a continuous fashion. Figure~\ref{fig:ivt} shows a counterexample
to this claim.

\section{Topos adapted to synthetic differential geometry}
\label{sect:smooth}

@@

* XXX phone call analogy
* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
* XXX allowing a switch of perspective
* XXX applications in mathematical practice
* XXX arbitrariness of the "standard axioms"
* XXX use the phrase "very rich and very deep" (or similarly) about set theory
* XXX A topos is more than its semantics
* XXX Check references everywhere (give good pointers)
* XXX internal/external

>>>>

%\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
%\author{Ingo Blechschmidt}
%\address{Universität Augsburg \\
%Institut für Mathematik \\
%Universitätsstr. 14 \\
%86159 Augsburg, Germany}
%\email{ingo.blechschmidt@math.uni-augsburg.de}

@@

\newextarrow{\xbigtoto}{{20}{20}{20}{20}}{\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\Zar}{Zar}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\effectively}{ef{}fectively\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

@@

%\begin{abstract}
%  FUTURE: abstract
%\end{abstract}

%\maketitle
\thispagestyle{empty}

\begin{center}
  \begin{minipage}{0.6\textwidth}
  \emph{To the anonymous referees:}
  \bigskip

  Thank you for your volunteer work. It is very much appreciated. I am looking
  forward to your comments.

  \hfill -- the anonymous author
  \end{minipage}
\end{center}
\bigskip

@@

\begin{figure}[b]
  \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
  \tikzstyle{title} = [fill=mypurple, text=white]

@@

\item We explicate how toposes provide distinct lenses through which the
usual mathematical objects of the standard topos can be viewed.
\end{enumerate}

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
mathematics. These are not entirely absolute; for instance, it is conceivable
that from the foundational crisis Brouwer's intuitionism would have emerged as
the main school of thought and we would now all reject the law of excluded
middle. Toposes allow us to explore alternatives to how history
has played out.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.\footnote{Olivia Caramello's research
program~\cite{caramello:tst} provides a further topos-theoretic way for
uncovering hidden relations, though not between objects but between mathematical
theories.}
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
We invite further research on these connections.

@@

\bigskip
\paragraph{Acknowledgments} We are grateful to Andrej Bauer, Martin
Brandenburg, Sina Hazratpour, Matthias Hutzler, Marc Nieper-Wißkirchen and
Alexander Oldenziel for invaluable discussions shaping this work, to Moritz
Laudahn, Matthias Hutzler and Milan Zerbin for their careful criticism of
earlier drafts and to Todd Lehman for preparing parts of
Figure~\ref{fig:landscape}. We thank the organizers and all participants of the
Mussomeli conference of the Italian Network for the Philosophy in Mathematics,
where this work was presented, for creating an exceptionally beautiful meeting. In
particular, we are grateful to Neil Barton, Danielle Macbeth, Gianluigi Oliveri
and Lorenzo Rossi for valuable comments.

@@

Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
so-called \emph{Kripke--Joyal translation rules}. For instance, the rule for
translating conjunction reads
\[ \E \models (\alpha \wedge \beta) \qquad\text{iff}\qquad
  \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
The remaining translation rules are more
involved~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
the case of a general topos~$\E$, but we will state them in the next sections
for several specific toposes. We refer to~``$\E \models \varphi$'' also as the
``external meaning of the internal statement~$\varphi$''.

@@

\subsection{The logic of toposes}\label{sect:logic-of-toposes}
By their definition as special kinds of
categories, toposes are merely algebraic structures not unlike groups or vector
spaces. Hence we need to argue why we picture toposes as mathematical universes
while we do not elevate other kinds of algebraic structures in the same way.
For us, this usage is justified by the following metatheorem:

@@

At this point in the text, all prerequisites for exploring toposes have been
introduced. The reader who wishes to develop, by explicit examples, intuition
for working internally to toposes is invited to skip ahead to
Section~\ref{sect:effective-topos}.

@@

\bigskip
\begin{center}
\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  % FUTURE: Countable axiom of choice
  \bottomrule
\end{tabular}
\end{center}
\bigskip

@@

\begin{ex}\textbf{``Any natural number is prime or not.''} Even without knowing what
a prime number is, one can safely judge this statement to be true in
the standard topos, since it is just an instance of the law of excluded middle.

By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
amounts to stating that there is a Turing machine which, when given a natural
number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
not. Such a Turing machine indeed exists -- writing such a program is often a
first exercise in programming courses. Hence the statement is also true in the
\effective topos, but for the nontrivial reason that primality can be
algorithmically tested.
\end{ex}

\begin{ex}\textbf{``There are infinitely many primes.''} A first-order formalization
of this statement is ``for any natural number~$n$, there is a prime
number~$p$ which is greater than~$n$'', and is known to be true in the standard
topos by any of the many proofs of this fact.

Its external meaning when interpreted in the \effective topos is that there exists
a Turing machine~$M$ which, when given a natural number~$n$ as input, terminates with a
prime number~$p > n$ as output. Such a Turing machine exists, hence the
statement is true in the \effective topos.\footnote{More precisely, the
machine~$M$ should also output the description of a Turing machine which
witnesses that~$p$ is prime and that~$p > n$. However, the statement~``$p$ is prime and~$p > n$''
is~$\neg\neg$-stable (even decidable), and for those statements witnesses are
redundant.}
\end{ex}

\begin{ex}\textbf{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

A rigorous proof that such a machine~$M$ does not exist reduces its assumed
existence to the undecidability of the halting problem.
\end{ex}

@@

\begin{ex}\textbf{``Any function~$\NN \to \NN$ is computable.''}
The preceding examples give the impression that what is true in the
\effective topos is solely a subset of what is true in the standard topos. The
example of this subsection, the so-called \emph{formal Church--Turing thesis},
shows that the relation between the two toposes is more nuanced.

@@

The usual proof sets up the halting function~$h : \NN \to \NN$, defined using
the case distinction
\[ h : n \mapsto \begin{cases}
  1, & \text{if the $n$-th Turing machine terminates}, \\
  0, & \text{if the $n$-th Turing machine does not terminate},
\end{cases} \]
and proceeds to show that~$h$ is not computable. However, in the \effective
topos, this definition does not give rise to a total function from~$\NN$
to~$\NN$. The actual domain is the subset of those natural numbers~$n$
for which the~$n$-th Turing machine
terminates or does not terminate. Assuming the law of excluded middle, this is
a trivial condition; but intuitionistically, it is not.
\end{ex}

\begin{ex}\textbf{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the sign and Heaviside functions
being prominent counterexamples. In the \effective topos, this statement is
true and independently due to
Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
straightforward (a textbook reference
is~\cite[Theorem~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
explanation is as follows.

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, when given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
such approximations. Hence a function such as the sign function, for which
even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
precision in the input~$x$, do not exist in the \effective topos.
\end{ex}

\begin{ex}\textbf{``Markov's principle holds.''} Markov's principle is the
following statement:
\begin{equation}\label{eq:markov}\tag{MP}
  \forall f \? \NN^\NN\_ \bigl((\neg\neg\exists n\?\NN\_ f(n) = 0)
  \Longrightarrow \exists n\?\NN\_ f(n) = 0\bigr).
\end{equation}
It is an instance of the law of double negation elimination and hence trivially
true in the standard topos, at least if we subscribe to classical logic on the
meta level. A useful consequence of Markov's principle is that Turing machines
which do not run forever (that is, which do \emph{not not} terminate) actually
terminate; this follows by applying Markov's principle to the function~$f : \NN
\to \NN$ where~$f(n)$ is zero or one depending on whether a given Turing
machine has terminated within~$n$ computational steps or not.

@@

As a consequence, the input given to machine~$S_F$ is entirely uninformative
and~$S_F$ cannot make direct computational use of it. But its existence ensures that an
\emph{unbounded search} will not fail (and hence succeed, by an appeal to Markov's
principle on the meta level): The machine~$S_F$ can simulate~$F$ to
compute the values~$f(0), f(1), f(2), \ldots$ in turn, and stop
with output~$n$ as soon as it determines that some function value~$f(n)$ is zero.
\end{ex}

\begin{ex}\textbf{``Heyting arithmetic is categorical.''} In addition to the standard
model~$\NN$, the standard topos contains uncountably many nonstandard models of
Peano arithmetic (at least if we assume a classical meta level). By a theorem
of Benno van den Berg and Jaap van Oosten, the situation
is quite different in the \effective topos~\cite{berg-oosten:arithmetic}:
\begin{enumerate}
\item Heyting arithmetic, the intuitionistic cousin of Peano arithmetic, has
exactly one model up to isomorphism, namely~$\NN$.
\item In fact, even the finitely axiomatizable subsystem of Heyting arithmetic
where the induction scheme is restricted to~$\Sigma_1$-formulas has exactly one
model up to isomorphism, again~$\NN$. As a consequence, Heyting arithmetic is
finitely axiomatizable.
\item Peano arithmetic is ``quasi-inconsistent'' in that it does not have any
models, for any model of Peano arithmetic would also be a model of Heyting
arithmetic, but the only model of Heyting arithmetic is~$\NN$ and~$\NN$ does
not validate the theorem ``any Turing machine terminates or does not
terminate'' of Peano arithmetic.
\end{enumerate}
As a consequence, Gödel's completeness theorem fails in the \effective topos:
In the \effective topos, Peano arithmetic is consistent (because it is
equiconsistent to Heyting arithmetic, which has a model) but does not have a
model.

The reason for~(1) is related to the fact that no nonstandard model of
Peano arithmetic in the standard topos is computable~\cite{tennenbaum:models}.
\end{ex}

@@

In the realizability topos constructed using infinite-time Turing machines, the full
law of excluded middle still fails, but some instances which are wrong in the
\effective topos do hold in this topos. For example, the instance ``any
function~$\NN \to \NN$ is the zero function or not'' does: Its external meaning
is that there is an infinite-time Turing machine~$M$ which, when given the
description of an infinite-time Turing machine~$F$ computing a function~$f :
\NN \to \NN$ as input, terminates (at some ordinal time step) with a correct
judgment of whether~$f$ is the zero function or not. Such a machine~$M$ indeed
exists: It simply has to simulate~$F$ on all inputs~$0,1,\ldots$ in order and
check whether one of the resulting function values is not zero. This
will require a transfinite amount of time (not least because simulating~$F$ on
just one input might require a transfinite amount of time), but as an
infinite-time Turing machine,~$M$ is capable of carrying out this procedure.

@@

@@

Just as the \effective topos provides a coherent setting for studying
computability using a naive language, the sheaf topos~$\Sh(X)$ provides a
coherent setting for studying continuous~$X$-indexed families of objects (sets,
numbers, functions) as if they were single objects.
% FUTURE: Explain better.
% FUTURE: History, sites, flexibility

@@

\subsection{Reifying continuous families of real numbers as single real numbers}
As detailed in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
actually a Turing machine computing arbitrarily-good consistent rational
approximations. A similarly drastic shift in meaning, though in an orthogonal
direction, occurs with~$\Sh(X)$. What~$\Sh(X)$ believes to be a (Dedekind) real
number~$a$ is actually a continuous family of real numbers on~$X$, that is, a
continuous function~$a : X \to \RR$~\cite[Corollary~D4.7.5]{johnstone:elephant}.

@@

\subsection{Reifying continuous families of real functions as single real functions}
Let~$(f_x)_{x \in X}$ be a continuous family of
continuous real-valued functions; that is, each of the individual
functions~$f_x : \RR \to \RR$ should be continuous and moreover the map~$\RR
\times X \to \RR, (a,x) \mapsto f_x(a)$ should be continuous.
From the point of view of~$\Sh(X)$, this family looks like a single continuous
function~$f : \RR \to \RR$.

@@

\begin{ex}
From these observations we can deduce that the intermediate value theorem of
undergraduate calculus does in general not hold in~$\Sh(X)$ and hence does not allow
for an intuitionistic proof. The intermediate value theorem states: ``If~$f : \RR \to \RR$
is a continuous function such that~$f(-1) < 0$ and~$f(1) > 0$, there exists a
number~$a$ such that~$f(a) = 0$.'' The external meaning of this statement is
that for any continuous family~$(f_x)_x$ of continuous functions with~$f_x(-1) <
0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible to pick zeros of
the family in a continuous fashion. Figure~\ref{fig:ivt} shows a counterexample
to this claim.

In contrast, the intermediate value theorem for monotone functions does have an
intuitionistic proof and hence applies in the internal universe of~$\Sh(X)$.
Thus for any continuous family~$(f_x)_x$ of continuous monotone functions
with~$f_x(-1) < 0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible
to pick zeros of the family in a continuous fashion.\footnote{While the
Kripke--Joyal translation of~``$\exists$'' is by definition \emph{local
existence}, one can show that the Kripke--Joyal translation of~``$\exists!$''
is \emph{unique existence on all opens}, in particular \emph{unique global
existence}. Because the conclusion of the intermediate value theorem for
monotone functions can be strengthened from ``has a zero'' to ``has a unique
zero'', this observation shows that the zeros can even globally be picked in a
continuous fashion.}
\end{ex}

\begin{ex}The fundamental theorem of algebra generally fails in~$\Sh(X)$, even
for quadratic polynomials. What~$\Sh(X)$ believes to be a (Dedekind)
complex number is externally a continuous function~$X \to \CC$. Let~$X$ be the
complex plane. Then the identity function~$\mathrm{id}_X$ is a single complex number from the
internal point of view of~$\Sh(X)$. The fundamental theorem of algebra would
predict~``$\Sh(X) \models \exists a\?\CC\_ a^2 - \mathrm{id}_X = 0$'',
hence that there is an open covering~$X = \bigcup_i U_i$ such that on each
open~$U_i$, there is a continuous function~$a : U_i \to \CC$ such that~$a(z)^2
- z = 0$ for all~$z \in U_i$. However, it is a basic fact of complex analysis
that such a function does not exist if~$0 \in U_i$.
\end{ex}

\begin{ex}The standard proof of Banach's fixed point theorem employs only
intuitionistic reasoning, hence applies internally to~$\Sh(X)$. Interpreting the
internal Banach fixed point theorem by the Kripke--Joyal translation rules
yields the statement that fixed points of continuous families of contractions
depend continuously on parameters.
\end{ex}

\section{Toposes adapted to synthetic differential geometry}
\label{sect:smooth}

@@

* XXX phone call analogy
FUTURE: quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory
* XXX A topos is more than its semantics
FUTURE: Check references everywhere (give good pointers)
